Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-10-25 manu-draft.qmd
more methods

### Seagrass monitoring and change

@@

From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite water quality conditions remaining relatively stable (Figure \[-@fig-seagrasschg\]a). Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to ta total of 30,137 acres (12,196 ha). Losses have been most pronounced in Old Tampa Bay (62%; 6,963acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599acres/647 ha loss). The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s. Similarly, coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss). Coverage in Lower Tampa Bay has remained stable, with only a 2% loss that is likely within the mapping error for the coverage estimates.

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD)

@@

EPC, FIM, Pinellas

@@

>>>>

### Seagrass change in Tampa Bay

@@

From 2016 to present, dramatic losses of seagrasses have been observed in Tampa Bay, despite water quality conditions remaining relatively stable (Figure [-@fig-seagrasschg]a). Total cover in Tampa Bay has decreased by 28% from the 2016 peak by 11,518 acres (4,661 ha) to ta total of 30,137 acres (12,196 ha). Losses have been most pronounced in Old Tampa Bay (62%; 6,963acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599acres/647 ha loss). The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s. Similarly, coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss). Coverage in Lower Tampa Bay has remained stable, with only a 2% loss that is likely within the mapping error for the coverage estimates.

### Seagrass data

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., , @SWFWMD23\]. These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December. The TBEP has used these maps to track progress towards achieving seagrass restoration goals as total cover in Tampa Bay. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure \[-@fig-seagrasschg\]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information, including abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow for inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Other sources of seagrass data are described in the next section.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s. These data provide the basis for regulatory assessments and compliance reporting for the nutrient TMDL in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements for physical parameters (e.g., salinity, temperature), where the latter includes measurements at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC). The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats. A stratified sample design is used to target multiple habitats where unique sites are sampled each month. We used data from the 21.3 meter center-bag nearshore seine that specifically targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present). These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021]. In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the beginning and end of the seine haul, and typically at the surface and bottom. Only measurements at the beginning of the seine haul and from the bottom were used. Seagrass data are also provided for each site, with information on species and cover. Sites with greater than 50% cover of seagrass were identified as "seagrass" sites and those less than 50% or bare sediment were identified as "no seagrass" sites for comparison with temperature and salinity measurements. Sites exclusively with macroalgae were not included in the analysis. All FIM data from FWC staff by request.

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PCDEM). Surface waters in Pinellas county have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting. Data were obtained by recquest to PCDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at \<https://wateratlas.usf.edu/\>). We focused primarily on OTB for the analysis of the PCDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments. Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained. Seagrass presence/absence is also recored at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program. This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data. Split-samples between the organizations are also compared to assess precision between different laboratories. As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.

@@

The first goal of this analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions from long-term climate change. Meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939. The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay. Annual averages for air temperature and cumulative annual precipitation were calculated. Additionally, the Standardized Precipitation Index (SPI) was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average (using the spi function from...).
